`infercnv` results exploration for `r params$sample_id`

### 2024-10-28 analyses/cell-type-wilms-tumor-06/notebook_template/06_cnv_infercnv_exploration.Rmd
update notebook

@@

srat_i3 <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_HMM-i3_", params$sample_id, "_reference-both.rds")))

@@

#### Plot UMAP of known Wilms tumor related alteration

As a reminder, we plot the label transfer from the fetal kidney atlas.

Immune and endothelial cells shouldn't show any CNV.

The stroma and fetal nephron compartments may contain normal cells with no CNV or cancer cells with low to high CNV profiles.

visualize_metadata(srat_i3, "fetal_kidney_predicted.compartment", "seurat_clusters")

#### Feature plot and repartition of the CNV per chromosome

This should allow a quick visual check that immune and endothelial cells do not expressed any CNV. 
If not, we should check if genes driving the CNV score are not key immune or endothelium genes.

The `umap` reduction visualization might allow the identification of cluster(s) of normal cells with no CNV.

```{r fig.width=10, fig.height=5, out.width='45%', out.height='45%'}
for(i in 1:22){
  print(visualize_feature(srat_i3, feature=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
}
```

####  CNV distribution in the Wilms tumor compartments

```{r fig.width=5, fig.height=5, out.width='40%', out.height='40%', warning=FALSE}
for(i in 1:22){
  print(visualize_density(srat_i3, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
}
```

#### DotPlot

The HMM-i3 prediction model with intra-patient reference (composed of immune and endothelium cells) perform really well in identifying Wilms tumor relevant CNV.
No false positive CNV are observed in the immune and endothelium compartment.

```{r fig.width=15, fig.height=5, out.width='100%'}

DotPlot(srat_i3, features =  colnames(srat_i3@meta.data)[grepl("has_cnv_chr", colnames(srat_i3@meta.data))], group.by = "fetal_kidney_predicted.compartment", assay = "SCT") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

#### CNV score

Here, we try to calculate a single CNV score and assess its potential in iditenfying cells with CNV versus normal cells without CNV.

We simply checked for each chromosome if the cell `has_cnv_chr`. 
Would the cell have one or more chromosome with CNV, the global `has_cnv_score` will be TRUE.
Else, the cell will have a `has_cnv_score` set to FALSE.

```{r fig.width=20, fig.height=5, out.width='100%', out.height='100%', warning=FALSE}
meta <- srat_i3@meta.data

meta$has_cnv_score <- rowSums(meta[,grepl("has_cnv_chr", colnames(meta))])
meta$has_cnv_score[meta$has_cnv_score > 1] <- "CNV"

# we know that some immune cells might have false positive CNV 
meta$has_cnv_score[meta$has_cnv_score %in% c(0,1)] <- "no CNV"

srat_i3 <- AddMetaData(srat_i3, meta$has_cnv_score, col.name =  "has_cnv_score")

We identified clusters with no/low CNV while the majority of the `Seurat` cluster present CNV on more than one chromosome.

visualize_metadata(srat_i3, "has_cnv_score", "seurat_clusters")

Immune and endothelial cells have a majority of cells with no/low CNV profile, which confirm their annotation as normal cells. 
The fetal nephron and stroma copartments are mainly composed of cancer cells with more than one chromosome affected by a CNV.

visualize_metadata(srat_i3, "has_cnv_score", "fetal_kidney_predicted.compartment")

@@

#### Feature plot and repartition of the CNV per chromosome

```{r fig.width=10, fig.height=5, out.width='45%', out.height='45%'}
for(i in 1:22){
  print(visualize_feature(srat_i6, feature=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
}
```

####  Distribution of CNV estimation in the Wilms tumor copartments

```{r fig.width=5, fig.height=5, out.width='40%', out.height='40%', warning=FALSE}
for(i in 1:22){
  print(visualize_density(srat_i6, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
}
```

#### DotPlot

The HMM-i3 prediction model with intra-patient reference (composed of immune and endothelium cells) perform really well in identifying Wilms tumor relevant CNV.
No false positive CNV are observed in the immune and endothelium compartment.

```{r fig.width=15, fig.height=5, out.width='100%'}

DotPlot(srat_i6, features =  colnames(srat_i6@meta.data)[grepl("has_cnv_chr", colnames(srat_i6@meta.data))], group.by = "fetal_kidney_predicted.compartment", assay = "SCT") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

#### CNV score

Finally, we try to calculate  single CNV score and assess its potential in iditenfying cells with CNV versus normal cells without CNV.

We simply checked for each chromosome if the cell `has_cnv_chr`. 
Would the cell have one or more chromosome with CNV, the global `has_cnv_score` will be TRUE.
Else, the cell will have a `has_cnv_score` set to FALSE.

```{r fig.width=20, fig.height=5, out.width='100%', out.height='100%', warning=FALSE}
meta <- srat_i6@meta.data

meta$has_cnv_score <- rowSums(meta[,grepl("has_cnv_chr", colnames(meta))])
meta$has_cnv_score[meta$has_cnv_score > 1] <- "CNV"

# we know that some immune cells might have false positive CNV 
meta$has_cnv_score[meta$has_cnv_score %in% c(0,1)] <- "no CNV"

srat_i6 <- AddMetaData(srat_i6, meta$has_cnv_score, col.name =  "has_cnv_score")

visualize_metadata(srat_i6, "has_cnv_score", "seurat_clusters")

visualize_metadata(srat_i6, "has_cnv_score", "fetal_kidney_predicted.compartment")

@@

```{r}
seurat_full <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_HMM-i3_", params$sample_id, "_reference-pull.rds")))
# subset non spike-in cells
keep_cells <- colnames(seurat_full)[!grepl("spike", colnames(seurat_full))]
seurat <- subset(seurat_full, cells = keep_cells)

meta <- seurat@meta.data

# merge the new info into the srat metadata
srat <- AddMetaData(srat, meta[,grepl("_cnv_chr", colnames(meta))], col.name = colnames(meta[grepl("_cnv_chr", colnames(meta))]))
```

#### Plot UMAP of known Wilms tumor related alteration

As a reminder, we plot the label transfer from the fetal kidney atlas.

visualize_metadata(srat, "fetal_kidney_predicted.compartment", "seurat_clusters")

#### Feature plot and repartition of the CNV per chromosome

This should allow a quick visual check that immune and endothelial cells do not expressed any CNV. If so, we should check if genes driving the CNV score are not key immune or endothelium genes.

Also, we should visualize if one area of the umap do not have CNV and could be identify as additional normal cell subset.

```{r fig.width=10, fig.height=5, out.width='45%', out.height='45%'}
for(i in 1:22){
  print(visualize_feature(srat, feature=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
}
```

####  Distribution of CNV estimation in the Wilms tumor copartments

```{r fig.width=5, fig.height=5, out.width='40%', out.height='40%', warning=FALSE}
for(i in 1:22){
  print(visualize_density(srat, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
}
```

#### DotPlot

The HMM-i3 prediction model with inter-patient reference (composed of immune and endothelium cells) perform quite well in identifying Wilms tumor relevant CNV.
However, sample variation induced false positive CNV such as CNV on chromosome 5 and 13 observed in the endothelium compartment.

```{r fig.width=15, fig.height=5, out.width='100%'}

DotPlot(srat, features =  colnames(srat@meta.data)[grepl("proportion_cnv_chr", colnames(srat@meta.data))], group.by = "fetal_kidney_predicted.compartment", assay = "SCT") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

#### CNV score

Finally, we try to calculate  single CNV score and assess its potential in iditenfying cells with CNV versus normal cells without CNV.

We simply checked for each chromosome if the cell `has_cnv_chr`. 
Would the cell have three or more chromosome with CNV, the global `has_cnv_score` will be TRUE.
Else, the cell will have a `has_cnv_score` set to FALSE.

Of note, as we know that the use of a ~universal inter-patient reference of normal cells can induce false positive CNV in some chromosomes, we decided to allow up to 2 CNV in normal cells.

```{r fig.width=20, fig.height=5, out.width='100%', out.height='100%', warning=FALSE}
meta <- srat@meta.data
meta$has_cnv_score <- rowSums(meta[,grepl("has_cnv_chr", colnames(meta))])
meta$has_cnv_score[meta$has_cnv_score > 2] <- "CNV"

# we know that some immune cells might have false positive CNV 
meta$has_cnv_score[meta$has_cnv_score %in% c(0,1, 2)] <- "no CNV"

srat <- AddMetaData(srat, meta$has_cnv_score, col.name =  "has_cnv_score")

visualize_metadata(srat, "has_cnv_score", "seurat_clusters")

visualize_metadata(srat, "has_cnv_score", "fetal_kidney_predicted.compartment")

@@

p1 <- SCpubr::do_DimPlot(srat, "has_cnv_score", plot.title = "inter-patient reference, HMM-i3", reduction = "umap")

p2 <- SCpubr::do_DimPlot(srat_i3, "has_cnv_score", plot.title = "intra-patient reference, HMM-i3", reduction = "umap")

p3 <- SCpubr::do_DimPlot(srat_i6, "has_cnv_score", plot.title = "intra-patient reference, HMM-i6", reduction = "umap")

>>>>

#### Wrapper function to explore `infercnv` HMM cnv prediction

The `wrapper_explore_hmm` take as input the `infercnv_obj` generated with `infercnv` HMM cnv predictions.
The wrapper allows the following steps and plots:

##### Feature plot and repartition of the CNV per chromosome

#####  Distribution of CNV estimation in the Wilms tumor copartments

##### DotPlot

##### CNV score

Finally, we try to calculate  single CNV score and assess its potential in identifying cells with CNV versus normal cells without CNV.

We simply checked for each chromosome if the cell `has_cnv_chr`. 
Would the cell have more than `cnv_threshold` chromosome with CNV, the global `has_cnv_score` will be TRUE.
Else, the cell will have a `has_cnv_score` set to FALSE.

wrapper_explore_hmm <- function(infercnv_obj, cnv_threshold = 1){

  p <- list()
  #### Plot UMAP of known Wilms tumor related alteration
  p[["compartment"]] <- visualize_metadata(infercnv_obj, "fetal_kidney_predicted.compartment", "seurat_clusters")
  #### Feature plot and repartition of the CNV per chromosome
  for(i in 1:22){
  tmp1 <- visualize_feature(infercnv_obj, feature=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment")
  tmp2 <- visualize_density(infercnv_obj, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment")
  p[[glue::glue("proportion_cnv_chr", i)]] <- tmp1 + tmp2 + plot_layout(ncol = 3, widths = c(1,2,2))
  }
  #### DotPlot
  p[["Dotplot"]] <- DotPlot(infercnv_obj, features =  colnames(infercnv_obj@meta.data)[grepl("has_cnv_chr", colnames(infercnv_obj@meta.data))], group.by = "fetal_kidney_predicted.compartment", assay = "SCT") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  #### CNV score
  meta <- infercnv_obj@meta.data
  meta$has_cnv_score <- rowSums(meta[,grepl("has_cnv_chr", colnames(meta))])
  meta$has_cnv_score[meta$has_cnv_score > cnv_threshold] <- "CNV"

  # we know that some immune cells might have false positive CNV 
  meta$has_cnv_score[meta$has_cnv_score %in% c(0:cnv_threshold)] <- "no CNV"
  infercnv_obj <- AddMetaData(infercnv_obj, meta$has_cnv_score, col.name =  "has_cnv_score") 
  p[["infercnv_obj"]] <- infercnv_obj

  p[["cnv_seurat"]] <- visualize_metadata(infercnv_obj, "has_cnv_score", "seurat_clusters")
  p[["cnv_compartment"]] <- visualize_metadata(infercnv_obj, "has_cnv_score", "fetal_kidney_predicted.compartment")

return(p)

}

@@

#### Load the `Seurat` object
srat_i3 <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_HMM-i3_", params$sample_id, "_reference-both.rds")))

@@

and explore the CNV results using the `wrapper_explore_hmm` fucntion.

```{r fig.width=16, fig.height=8, out.width='100%', out.height='100%', warning=FALSE}
p <- list()
p <- wrapper_explore_hmm(srat_i3, cnv_threshold = 1)
p
```

@@

```{r fig.width=16, fig.height=8, out.width='100%', out.height='100%', warning=FALSE}
q <- list()
q <- wrapper_explore_hmm(srat_i6, cnv_threshold = 1)
q
```

@@

```{r}
seurat_full <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_HMM-i3_", params$sample_id, "_reference-pull.rds")))
# subset non spike-in cells
keep_cells <- colnames(srat)
seurat <- subset(seurat_full, cells = keep_cells)

meta <- seurat@meta.data[colnames(srat),]

# merge the new info into the srat metadata
srat_pull <- AddMetaData(srat, meta[,grepl("_cnv_chr", colnames(meta))], col.name = colnames(meta[grepl("_cnv_chr", colnames(meta))]))
```

```{r fig.width=16, fig.height=8, out.width='100%', out.height='100%', warning=FALSE}
pull <- list()
pull <- wrapper_explore_hmm(srat_pull, cnv_threshold = 3)
pull
```

@@

p1 <- SCpubr::do_DimPlot(pull$infercnv_obj, "has_cnv_score", plot.title = "inter-patient reference, HMM-i3", reduction = "umap")

p2 <- SCpubr::do_DimPlot(p$infercnv_obj, "has_cnv_score", plot.title = "intra-patient reference, HMM-i3", reduction = "umap")

p3 <- SCpubr::do_DimPlot(q$infercnv_obj, "has_cnv_score", plot.title = "intra-patient reference, HMM-i6", reduction = "umap")
